Location and Scale-Invariant Power Transformations for Transforming Data to

### 2025-04-09 manuscript_rmarkdown/manuscript.Rmd
Updated machine learning results.

We used 231 datasets containing at least one numeric feature from the Penn Machine Learning Benchmarks collection [@Romano2022-gq].
In this collection, 114 datasets correspond to regression tasks and 117 datasets
to classification tasks. Using the familiar auto-machine learning library [@Zwanenburg2021-so] (version 1.5.0),
each dataset was used to train a model for each of 32 process configurations.
Each process configuration specifies the learner (generalised linear model, L1-regularised linear models (Lasso), gradient boosted linear model, or
random forest), transformation method (none, conventional Yeo-Johnson, robust
invariant Yeo-Johnson, robust invariant Yeo-Johnson with empirical central
normality test (rejecting transformations with $p \leq 0.01$), and normalisation method (none,
$z$-standardisation), yielding 32 distinct configurations.
Before each experiment, each dataset was randomly split into a training (70%)
and holdout test (30%) set five times. Thus, a total of 27720 models were
created. Each model was then evaluated using the holdout test set using one of
two metrics, i.e. the root relative squared error (RRSE) for regression tasks and the
area under the receiver operating characteristic curve (AUC) for classification tasks.

@@

To estimate the marginal effects of process parameters, including transformation method, 
we first fit a regression random forest (ranger package [@Wright2017-rf] version 0.16.0): 2000 trees, 
node size 2, other hyperparameters default) with process parameters and task difficulty as predictors and normalised rank as response variable.
The estimated marginal effects are shown in Figure \ref{fig:marginal-effect-plot}.
On the scale of normalised ranks ($[0.0, 1.0]$), the overall estimated marginal effect of using a random forest instead of generalised linear model was $0.272$.
The overall marginal effect of using z-standardisation to normalise features was $0.008$.
Transformation methods had the following marginal effects:
$0.003$ for using conventional Yeo-Johnson transformation instead of no transformation;
$-0.007$ for using robust invariant Yeo-Johnson transformation instead of no transformation; 
$-0.009$ for using robust invariant Yeo-Johnson transformation with empirical central normality test instead of no transformation; 
$-0.010$ for using robust invariant Yeo-Johnson transformation instead of conventional Yeo-Johnson transformation;
and $-0.012$ for using using robust invariant Yeo-Johnson transformation with empirical central normality test instead of conventional Yeo-Johnson transformation.

```{r marginal-effect-plot, echo=FALSE, fig.cap=cap, fig.height=7.5, warning=FALSE}
cap <- paste0(
  "Estimated marginal effect of learners, normalisation and transformation methods on ranked model performance scores in 18560 machine learning experiments on 232 datasets. ",
  "The top-left panel shows the marginal effect of learners, i.e. random forests and generalised linear models (GLM). ",
  "Random forests outperform GLM models for all task difficulties. ",
  "The top-right panel shows the marginal effect of feature normalisation methods, i.e. no normalisation and z-standardisation. ",
  "z-standardisation is generally beneficial, but the estimated effect is neglible. ",
  "The bottom panel shows the marginal effects of different transformation methods, split by learner and normalisation method. ",
  "There is no consistent behaviour, and estimated effects are neglible. ",
  "Note that the ranges of the $x$-axes of the three main panels differ. ",
  "ECNT: empirical central normality test."
)

@@

Robust location- and scale-invariant transformations are a suitable replacement for their conventional counterparts.
They demonstrated robustness against outliers and prevent inaccurate transformations and potential numerical issues due to location and scale of the distribution of a feature.
This is particularly relevant for automated data processing, where such issues may go unnoticed.
Compared to non-robust location- and scale-invariant transformations, real-world examples with outliers showed higher residual errors of transformed features. 
Robust transformations seek to minimise residual errors for the central part of the distribution, instead of the entire distribution, including outliers.
The empirical central normality test indicated that robust transformations are better able to achieve central normality in the presence of outliers.
However, in a machine learning experiment of 232 real-world datasets that contained at least one numeric feature, we did not find a meaningful benefit -- nor detriment -- to model performance for location- and scale-invariant power transformations.
One reason may be that numeric features with large location shifts ($|\mu| > 1000.0$) were uncommon.
Of the 4886 numeric features in the 232 datasets, 266 (5%) features in 34 datasets had large location shifts, of which 200 appeared in just 2 datasets.
For the latter two datasets, the transformation method did not show significant difference between groups (Friedman test; $p > 0.05$).
```{r include = FALSE, eval = TRUE}

>>>>

We used 231 datasets containing at least one numeric feature from the Penn Machine Learning Benchmarks collection [@Romano2022-gq].
In this collection, 114 datasets correspond to regression tasks and 117 datasets
to classification tasks. Using the familiar auto-machine learning library [@Zwanenburg2021-so] (version 1.5.0),
each dataset was used to train a model for each of 32 process configurations.
Each process configuration specifies the learner (generalised linear model, L1-regularised linear models (Lasso), gradient boosted linear model, or
random forest), transformation method (none, conventional Yeo-Johnson, robust
invariant Yeo-Johnson, robust invariant Yeo-Johnson with empirical central
normality test (rejecting transformations with $p \leq 0.01$), and normalisation method (none,
$z$-standardisation), yielding 32 distinct configurations.
Before each experiment, each dataset was randomly split into a training (70%)
and holdout test (30%) set five times. Thus, a total of 36960 models were
created. Each model was then evaluated using the holdout test set using one of
two metrics, i.e. the root relative squared error (RRSE) for regression tasks and the
area under the receiver operating characteristic curve (AUC) for classification tasks.

@@

To estimate the marginal effects of process parameters, including transformation method, 
we first fit a regression random forest (ranger package [@Wright2017-rf] version 0.16.0): 2000 trees, 
node size 2, other hyperparameters default) with process parameters and task difficulty as predictors and normalised rank as response variable.
The estimated marginal effects are shown in Figure \ref{fig:marginal-effect-plot}.
On the scale of normalised ranks ($[0.0, 1.0]$), only the random forest performed better than the average of all learners (rank difference $0.173$).
The overall marginal effect of using z-standardisation to normalise features was $0.011$.
Transformation methods had the following marginal effects:
$-0.001$ for using conventional Yeo-Johnson transformation instead of no transformation;
$-0.012$ for using robust invariant Yeo-Johnson transformation instead of no transformation; 
$-0.012$ for using robust invariant Yeo-Johnson transformation with empirical central normality test instead of no transformation; 
$-0.012$ for using robust invariant Yeo-Johnson transformation instead of conventional Yeo-Johnson transformation;
and $-0.012$ for using using robust invariant Yeo-Johnson transformation with empirical central normality test instead of conventional Yeo-Johnson transformation.

```{r marginal-effect-plot, echo=FALSE, fig.cap=cap, fig.height=7.5, warning=FALSE}
cap <- paste0(
  "Estimated marginal effect of learners, normalisation and transformation methods on ranked model performance scores in 36960 machine learning experiments on 231 datasets. ",
  "The four top-left panels shows the marginal effect of learners vs. the average, i.e. generalised linear models (GLM), L1-regularised LM (Lasso), gradient-boosted LM and random forest. ",
  "Random forests outperform the average. ",
  "The top-right panel shows the marginal effect of feature normalisation methods, i.e. no normalisation and z-standardisation. ",
  "z-standardisation is generally beneficial, but the estimated effect is marginal. ",
  "The bottom panel shows the marginal effects of different transformation methods, split by normalisation method. ",
  "The estimated effects are similar in size to the effect of normalisation and marginal. ",
  "Note that the ranges of the $x$-axes of the three main panels differ. ",
  "ECNT: empirical central normality test."
)

@@

Robust location- and scale-invariant transformations are a suitable replacement for their conventional counterparts.
They demonstrated robustness against outliers and prevent inaccurate transformations and potential numerical issues due to location and scale of the distribution of a feature.
This is particularly relevant for automated data processing, where such issues may go unnoticed.
Compared to non-robust location- and scale-invariant transformations, real-world examples with outliers showed higher residual errors of transformed features. 
Robust transformations seek to minimise residual errors for the central part of the distribution, instead of the entire distribution, including outliers.
The empirical central normality test indicated that robust transformations are better able to achieve central normality in the presence of outliers.
However, in a machine learning experiment of 231 real-world datasets that contained at least one numeric feature, we did not find a meaningful benefit -- nor detriment -- to model performance for location- and scale-invariant power transformations.
One reason may be that numeric features with large location shifts ($|\mu| > 1000.0$) were uncommon.
Of the 4886 numeric features in the 231 datasets, 266 (5%) features in 34 datasets had large location shifts, of which 200 appeared in just 2 datasets.
For the latter two datasets, the transformation method did not show significant difference between groups (Friedman test; $p > 0.05$).
```{r include = FALSE, eval = TRUE}
